Road accidents remain one of the leading causes of death and serious injury 
worldwide. According to the World Health Organization, approximately 1.19 million 
people die each year as a result of road traffic crashes, with tens of millions 
more sustaining non-fatal injuries~\cite{who2023roadsafety}. Beyond the human 
toll, road accidents impose substantial economic burdens on healthcare systems, 
emergency response services, and transportation infrastructure. A critical 
operational challenge within road safety management is the timely and accurate 
assessment of accident severity, that is, whether a reported incident is likely 
to result in a slight, serious, or fatal injury. Reliable severity prediction has 
direct practical value: it can inform how emergency responders prioritize 
dispatch, how traffic authorities allocate road safety budgets, and how 
policymakers identify high-risk conditions that warrant targeted intervention.

Machine learning has emerged as a promising approach for this prediction task. 
Several studies have applied classification algorithms to road accident datasets 
with varying degrees of success. Ahmed et al.~\cite{ahmed2023explainable} 
demonstrated that ensemble models such as Random Forest could identify key 
contributing factors, including road category and number of vehicles involved, 
while achieving strong predictive performance on accident severity data. Obasi 
and Benson~\cite{obasi2023evaluating} conducted a comparative analysis using ten 
years of UK traffic accident data (2005--2014), evaluating Naive Bayes, Logistic 
Regression, Random Forest, and Artificial Neural Networks, and reported overall 
prediction accuracies of up to 87\%. Similarly, Ardakani et 
al.~\cite{ardakani2023road} applied decision trees, random forest, logistic 
regression, and Naive Bayes to the same UK dataset, evaluating performance across 
multiple prediction targets including accident severity, casualty count, and 
vehicle count. More recently, Mohamad et al.~\cite{mohamad2025thailand} conducted 
a multi-model comparison across eight classifiers on a large-scale Thai accident 
dataset, finding that Random Forest consistently outperformed simpler models in 
separating fatal from non-fatal outcomes.

Despite this growing body of work, a recurring and underaddressed limitation is 
the treatment of class imbalance. In real-world accident datasets, severe and 
fatal incidents are vastly outnumbered by minor ones. In the UK dataset used in 
this study, approximately 85\% of records are classified as slight injuries while 
fatal accidents account for barely 1.3\% of all cases. When class imbalance is 
left unaddressed, classifiers tend to inflate overall accuracy by defaulting to 
the majority class, while performing poorly on the minority classes that carry 
the greatest safety consequences. Fiorentini and Losa~\cite{fiorentini2020imbalance} 
reviewed dozens of crash severity prediction studies and found that most 
overlooked the class imbalance problem entirely, resulting in classifiers that 
performed well on majority classes while failing to detect the more severe 
minority cases. Yet few comparative studies evaluate how classifiers perform both 
with and without explicit imbalance-handling strategies within the same 
experimental framework.

This study addresses that gap directly. We conduct a systematic comparative 
analysis of five widely-used classification algorithms, namely Logistic 
Regression, Naive Bayes, Decision Tree, Random Forest, and Support Vector 
Machine, applied to the UK Road Accident dataset (2005--2014). The study plans 
to evaluate each model under two conditions: trained on the original imbalanced 
data, and trained with imbalance-handling strategies applied. This design allows 
us to isolate and quantify how much imbalance handling improves each classifier's 
ability to detect serious and fatal accidents, which are the cases that matter 
most for emergency response and road safety policy. By providing this direct 
comparison on a large, well-documented public dataset, this study offers 
practitioners clearer guidance on which modeling strategies are genuinely 
reliable for safety-critical severity classification, rather than which simply 
achieve high accuracy on easy majority-class predictions.